Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-10 csf-2017/paper.tex
Finish the discussion about invoking the adversary.

\paragraph{Program startup}
When our code starts executing, we assume that the language runtime initializes
the memory as follows: a contiguous array of memory is reserved for the stack,
for which we receive a stack pointer in a register that we reserve for this
purpose \dominique{give this register a special name like sp?}. This stack
pointer must be local and it must carry $\rwlx$ permission. It is crucial for
our approach that the stack is the only part of memory for which there will ever
be capabilities with $\rwlx$ or $\rwl$ capabilities. We expect the language
runtime to enforce this by not handing out such capabilities for other parts of
memory in any situation (i.e. not through malloc, not during loading or linking
etc.). Additionally, in our examples, we typically also assume some static parts
of memory where instructions are stored or static data. Another part of memory
(called the heap) is initially governed by malloc and at program startup, no
other code has capabilities for this memory. Malloc will hand out capabilities
for allocated regions as requested, but it will never hand out capabilities with
write-local capability (i.e. no $\rwlx$ or $\rwl$ permissions). For simplicity,
we assume that memory allocated through malloc cannot be freed.

\paragraph{Invoking the adversary}
During execution, we use the stack as normal: we may push values onto it and pop
them back off (e.g. for register spilling etc.). When we invoke the adversary,
we need to take great care to only provide him with the authority we want him to
have. Specifically, we clear all the registers, except those that contain
arguments the adversary is intended to have access to. We leave a copy of the
stack pointer in the appropriate register, but only after we have used the
$\subseg{}{}{}$ to shrink its authority to just the part of the stack that we
are not using. In one of the registers, we will also provide a return pointer.
We will go into constructing the return pointer further below.

@@

\paragraph{Returning from the adversary}
So return pointers must be passed as local capabilities. But what should their
permissions be, what memory should they point to and what should that memory
(let's call it the activation record) contain? Let us answer the last question
first by considering what should happen when the adversary jumps to a return
pointer and what capabilities we need for achieving that. Essentially, the
program counter should be restored to the code after the call into the
adversary, so the activation record should store this program counter.
Additionally, the stack pointer should also be restored to its original value.
Since the adversary has a more restricted authority over the stack than the code
making the call, we cannot hope to reconstruct the return pointer from the stack
pointer owned by the adversary. Instead, the old stack pointer should also be
stored as part of the activation record.

\begin{figure}
  \begin{tikzpicture}[scale=.7, every node={scale=.7}]
    \draw[fill=gray!20,draw=none] (0,4) rectangle (6,3.5);
    \draw[fill=gray!20,draw=none] (0,0.5) rectangle (6,1);
    \draw (0,0.5) -- (0,4);
    \draw (6,0.5) -- (6,4);
    \draw (0,3.5) rectangle (6,3) node[pos=.5] {\footnotesize previous program counter};
    \draw (0,3) rectangle (6,2.5) node[pos=.5] {\footnotesize previous stack pointer};
    \draw (0,2.5) rectangle (6,1) node[pos=.5] {\footnotesize restore instructions};
    \draw (-3,2.25) node (rp) {return pointer};
    \draw[->] (rp.east) -- (-.2,2.25);
  \end{tikzpicture}

@@

It's important that neither of these capabilities are accessible to the
adversary. In other words, we need to provide the adversary a capability that he
can jump to but not read from, i.e. an enter capability. To make this work, we
construct the activation record as depicted in Figure~\ref{fig:activ-rec-struct}.

* return pointer as enter pointer to activation
record on the stack. * activation record contains previous pc and stack pointer

\paragraph{Invoking an adversary callback}
* must be global

>>>>

\textbf{Program startup}
When our code starts executing, we assume that the language runtime initializes
the memory as follows: a contiguous array of memory is reserved for the stack,
for which we receive a stack pointer in a register that we reserve for this
purpose \dominique{give this register a special name like sp?}. This stack
pointer must be local and it must carry $\rwlx$ permission. It is crucial for
our approach that the stack is the only part of memory for which there will ever
be capabilities with $\rwlx$ or $\rwl$ capabilities. We expect the language
runtime to enforce this by not handing out such capabilities for other parts of
memory in any situation (i.e. not through malloc, not during loading or linking
etc.). Additionally, in our examples, we typically also assume some static parts
of memory where instructions are stored or static data. Another part of memory
(called the heap) is initially governed by malloc and at program startup, no
other code has capabilities for this memory. Malloc will hand out capabilities
for allocated regions as requested, but it will never hand out capabilities with
write-local capability (i.e. no $\rwlx$ or $\rwl$ permissions). For simplicity,
we assume that memory allocated through malloc cannot be freed.

\textbf{Invoking the adversary}
During execution, we use the stack as normal: we may push values onto it and pop
them back off (e.g. for register spilling etc.). When we invoke the adversary,
we need to take great care to only provide him with the authority we want him to
have. Specifically, we clear all the registers, except those that contain
arguments the adversary is intended to have access to. We leave a copy of the
stack pointer in the appropriate register, but only after we have used the
$\subseg{}{}{}$ to shrink its authority to just the part of the stack that we
are not using. In one of the registers, we will also provide a return pointer.
We will go into constructing the return pointer further below.

@@

\textbf{Returning from the adversary}
So return pointers must be passed as local capabilities. But what should their
permissions be, what memory should they point to and what should that memory
(let's call it the activation record) contain? Let us answer the last question
first by considering what should happen when the adversary jumps to a return
pointer and what capabilities we need for achieving that. Essentially, the
program counter should be restored to the code after the call into the
adversary, so the activation record should store this program counter.
Additionally, the stack pointer should also be restored to its original value.
Since the adversary has a more restricted authority over the stack than the code
making the call, we cannot hope to reconstruct the return pointer from the stack
pointer owned by the adversary. Instead, the old stack pointer should also be
stored as part of the activation record.

\begin{figure}
  \begin{tikzpicture}[scale=.7, every node={scale=.7}]
    \draw[fill=gray!20,draw=none] (0,4) rectangle (6,3.5);
    \draw[fill=gray!20,draw=none] (0,0.5) rectangle (6,1);
    \draw[draw=gray!80] (0,0.5) -- (0,4);
    \draw[draw=gray!80] (6,0.5) -- (6,4);
    \draw (0,3.5) rectangle (6,3) node[pos=.5] {\footnotesize previous program counter};
    \draw (0,3) rectangle (6,2.5) node[pos=.5] {\footnotesize previous stack pointer};
    \draw (0,2.5) rectangle (6,1) node[pos=.5] {\footnotesize restore instructions};
    \draw (-3,2.25) node (rp) {return pointer};
    \draw[->] (rp.east) -- node[above] {$\entry$}(-.2,2.25);
  \end{tikzpicture}

@@

It's important that neither of these capabilities are accessible to the
adversary. In other words, we need to provide the adversary a capability that he
can jump to but not read from, i.e. an enter capability. To make this work, we
construct the activation record as depicted in
Figure~\ref{fig:activ-rec-struct}. The return pointer is an $\entry$ capability
pointing with authority over a data structure containing the previous return and
stack pointer as well as a number of restore instructions. The current address
points to those instructions, so that, upon invocation, they can load the old
stack pointer and program counter into the register file through the authority
of the pc. The return pointer is given to the adversary as an enter pointer so
that the adversary cannot get hold of the contents of the data structure, but
after invocation, its permission is updated to $\exec$, so that the activation
record's contents become available to the restore instructions.

The final question that remains is: where should we construct this return
pointer? The attentive reader will see that there is only one possibility: since
the activation record contains the old stack pointer, which is necessarily
local, the activation record can only be constructed in a part of memory where
we have write-local access, i.e. on the stack. Note that this means we will be
placing and executing instructions on the stack, i.e. it will not just contain
code pointers and data.  This means that this discipline should be combined with
protection against stack smashing attacks (i.e. buffer overflows on the stack overwriting activation
records' contents), but capability machines' fine-grained memory protection
luckily is well-suited for this.

\textbf{Invoking an adversary callback}
* must be global
